---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
activity<-read.csv("activity.csv")

##remove N
activity_cc<-as.data.frame(activity[complete.cases(activity),])
dates<-unique(as.Date(activity_cc$date))
days<-split(activity_cc,as.factor(activity_cc$date))
```

## What is mean total number of steps taken per day?
```{r numSteps}
##do lapply with a function that does the some of the data frame
doSum<-function(day) {
  df<-as.data.frame(day)
  sum(df[,1])
}

##breaking the data into days
dates<-unique(as.Date(activity_cc$date))
days<-split(activity_cc,as.factor(activity_cc$date))

##creating the daily total for the steps
dailySteps<-as.numeric(lapply(days,doSum))
dailySteps<-dailySteps[dailySteps>0]  ##ignoring days where there are no steps as first and last day is NA
##Histogram of total daily steps
hist(dailySteps)

##summary including the mean and median
summary(dailySteps)


```



## What is the average daily activity pattern?

I will spread the complete data set so that each interval is a column (fat data.)  I will then create the mean of the columns
to obtain the mean steps per interval.  I will get the time intervals from the names of this dataset and then plot the means vs the times

```{r dailyActivity}
require(tidyr)
activity_int<-spread(activity_cc,interval,steps)

sumcol<-subset(activity_int, select = -date)
cmeans<-colMeans(sumcol)
ctimes<-names(activity_int)
ctimes<-ctimes[2:289]
plot(x=ctimes,y=cmeans,type="l",xlab = "Time (in 5 min intervals",ylab = "Mean Steps", main="Daily Activity Pattern")
```



## Imputing missing values
NOTE: My method for creating the missing values is taking the average for each interval using the dataset with only complete clases (x2
)  I then get a list of incomplete rows and for each empty value I replace it with the mean value for that 5 minute interval.
Calculate the rows that need to be replaced
```{R calcRep}
replace<-which(is.na(activity$steps))
```

Calculate the number of rows by getting the length
```{R numRep}
length(replace)
```

Do the replacement
```{R replace}
## for each item in x if is.na(Steps) replace with
## cmeans[row]
for(i in replace) {
    row<-i %% 288 + 1  ##each day has 288 intervals so I figure out what interval it is using the modulo of 288 +1 (since its base 1)
    rep<-as.integer(cmeans[row])
    activity[i,1]<- rep
}
```

Create the histogram of data with NA's replaced
```{R dailySteps2}
dates<-unique(as.Date(activity$date))
days<-split(activity,as.factor(activity$date))
dailySteps<-as.numeric(lapply(days,doSum))
dailySteps<-dailySteps[dailySteps>0]
hist(dailySteps)
```

The summary of daily steps where NA is replaced

```{r summary}
summary(dailySteps)
```


## Are there differences in activity patterns between weekdays and weekends?

Process the updated data (where NA is replace) to create separate datasets for weekends and weekdays and create timeseries graphs
Note that the timeseries graphs appear to be as one would expace....more steps earlier on the weekdays with steps decreasing earlier.  On the weekends steps jump up a little later but there are more steps during the day and longer into the evening.
```{r activityCompare}
##create the weekday and weekend column
dows<-weekdays(as.Date(activity$date))
dows[dows=="Sunday" | dows=="Saturday"]<-"Weekend"
dows[!(dows=="Weekend")]<-"Weekday"

##bind it to the updated data and then splt using the weekend/weekday column as a factor
x_wd<-cbind(activity,dows)
days2<-split(x_wd,as.factor(x_wd$dows))
wdVal<-days2[[1]]
weVal<-days2[[2]]

##spread the data and calculate the means as was done earlier
wdVal2<-subset(wdVal,select = -dows)
wd2<-spread(wdVal2,interval,steps)

sumcol<-subset(wd2, select = -date)
cmeans<-colMeans(sumcol)
ctimes<-names(wd2)
ctimes<-ctimes[2:289]

##setup and create the first plot
par(mfcol=c(2,1))
par(mar=c( 4.1, 4.1, 1.1, 2.1))
plot(x=ctimes,y=cmeans,type="l",xlab = "Time (in 5 min intervals",ylab = "Mean Steps", main = "Weekday Steps",cex.main=.75,cex.axis=0.75,cex.lab=.75)

##do the same for the weekend data
weVal2<-subset(weVal,select = -dows)
we2<-spread(weVal2,interval,steps)

sumcol<-subset(we2, select = -date)
cmeans<-colMeans(sumcol)
ctimes<-names(we2)
ctimes<-ctimes[2:289]
plot(x=ctimes,y=cmeans,type="l",xlab = "Time (in 5 min intervals",ylab = "Mean Steps", main = "Weekend Steps",cex.main=.75,cex.axis=0.75,cex.lab=.75)

```
