---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
x<-read.csv("activity.csv")

##remove N
x2<-as.data.frame(x[complete.cases(x),])
dates<-unique(as.Date(x2$date))
days<-split(x2,as.factor(x2$date))
```

## What is mean total number of steps taken per day?
```{r numSteps}
##do lapply with a function that does the some of the data frame
doSum<-function(day) {
  df<-as.data.frame(day)
  sum(df[,1])
}

##breaking the data into days
dates<-unique(as.Date(x2$date))
days<-split(x2,as.factor(x2$date))

##creating the daily total for the steps
dailySteps<-as.numeric(lapply(days,doSum))
dailySteps<-dailySteps[dailySteps>0]  ##ignoring days where there are no steps as first and last day is NA
##Histogram of total daily steps
hist(dailySteps)

##summary including the mean and median
summary(dailySteps)


```



## What is the average daily activity pattern?

```{r dailyActivity}
require(tidyr)
x4<-spread(x2,interval,steps)

sumcol<-subset(x4, select = -date)
cmeans<-colMeans(sumcol)
ctimes<-names(x4)
ctimes<-ctimes[2:289]
plot(x=ctimes,y=cmeans,type="l",xlab = "Time (in 5 min intervals",ylab = "Mean Steps")
```



## Imputing missing values
NOTE: My method for creating the missing values is taking the average for each interval using the dataset with only complete clases (x2
)  I then get a list of incomplete rows and for each empty value I replace it with the mean value for that 5 minute interval.
```{R dailySteps2}
x4<-spread(x2,interval,steps)

sumcol<-subset(x4, select = -date)
cmeans<-colMeans(sumcol)

replace<-which(is.na(x$steps))

print(length(replace))
## for each item in x if is.na(Steps) replace with
## cmeans[row]
for(i in replace) {
    row<-i %% 288 + 1  ##each day has 288 intervals so I figure out what interval it is using the modulo of 288 +1 (since its base 1)
    rep<-as.integer(cmeans[row])
    x[i,1]<- rep
}
dates<-unique(as.Date(x$date))
days<-split(x,as.factor(x$date))
dailySteps<-as.numeric(lapply(days,doSum))
dailySteps<-dailySteps[dailySteps>0]
hist(dailySteps)
summary(dailySteps)
```


## Are there differences in activity patterns between weekdays and weekends?

```{r activityCompare}
dows<-weekdays(as.Date(x$date))
dows[dows=="Sunday" | dows=="Saturday"]<-"Weekend"
dows[!(dows=="Weekend")]<-"Weekday"
x_wd<-cbind(x,dows)
days2<-split(x_wd,as.factor(x_wd$dows))
wdVal<-days2[[1]]
weVal<-days2[[2]]
par(mfcol=c(2,1))
par(mar=c( 2.1, 4.1, 2.1, 2.1))
wdVal2<-subset(wdVal,select = -dows)
wd2<-spread(wdVal2,interval,steps)

sumcol<-subset(wd2, select = -date)
cmeans<-colMeans(sumcol)
ctimes<-names(wd2)
ctimes<-ctimes[2:289]
plot(x=ctimes,y=cmeans,type="l",xlab = "Time (in 5 min intervals",ylab = "Mean Steps", main = "Weekday Steps")
weVal2<-subset(weVal,select = -dows)
we2<-spread(weVal2,interval,steps)

sumcol<-subset(we2, select = -date)
cmeans<-colMeans(sumcol)
ctimes<-names(we2)
ctimes<-ctimes[2:289]
plot(x=ctimes,y=cmeans,type="l",xlab = "Time (in 5 min intervals",ylab = "Mean Steps", main = "Weekend Steps")

```
